---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
```

# dsir

<!-- badges: start -->
<!-- badges: end -->

### Status: Experimental testing only

The `dsir` package provides an easy way for researchers and governments to obtain country data on digital sequence information (genetic sequence data) submitted to [NCBI databases](https://www.ncbi.nlm.nih.gov/guide/all/) such as [GenBank](https://www.ncbi.nlm.nih.gov/genbank/).

Digital Sequence Information (DSI) is a placeholder term for genetic sequence data in international policy debates. The governance of DSI has become a major focus of debate under the [United Nations Convention on Biological Diversity and its Nagoya Protocol](https://www.cbd.int/dsi-gr/) as well as the [Plant Treaty](http://www.fao.org/cgrfa/topics/digital-sequence-information/en/), the ongoing negotiations of the [new treaty on marine biodiversity](https://www.un.org/bbnj/) (Biodiversity Beyond National Jurisdiction or BBNJ) under UNCLOS and the sharing of virus materials under the [PIP Framework](https://www.who.int/influenza/pip/en/) at the World Health Organisation (WHO). The pandemic has sharpened attention to the governance of sequence data.

One challenge for governments in these debates, notably from developing countries, is identifying up to date data on sequence submissions that originate in samples from their country to inform evidence based discussion. This is the challenge that `dsir` aims to help solve. 

Existing tools for sequence data focus on submitting and accessing sequence listings. In contrast, for policy discussions what is mainly of interest is the `metadata` about the sequences. What is needed is a slimmed down approach that focuses on metadata... while allowing sequence data to be accessed as needed. 

This package provides the following:

- A small set of R functions to obtain up to date NCBI esummary (metadata) on sequences from countries 
- A set of country datasets to allow researchers to explore the data in detail.

`dsir` provides simplified wrappers for some of the functions from the [ROpenSci](https://ropensci.org/) [rentrez](https://docs.ropensci.org/rentrez/) package created by David Winter and collaborators. `rentrez` provides access to the NCBI [E-Utils API](https://www.ncbi.nlm.nih.gov/books/NBK25500/) for remotely accessing NCBI databases. The credit for the underlying code that powers `dsir` therefore belongs with David Winter and collaborators. You can visit `rentrez` [here]([https://github.com/ropensci/rentrez](https://github.com/ropensci/rentrez)) and give it some love by leaving a star.

`dsir` is designed to be easy to use for people who are new to R and to facilitate the automated retrieval and monitoring of sequence data originating from a country. However, the underlying `rentrez` package provides a powerful set of tools for the exploration and retrieval of sequence related data from the network of NCBI databases. More advanced R users will want to engage with the rich function set provided by the `rentrez` package.


## Installation

If you have arrived here and you are completely new to the R programming language, do not fear. You need to start by installing R and RStudio on your computer following the instructions in Garrett Grolemund's excellent and free [Hands On Programming with R](https://rstudio-education.github.io/hopr/). Follow Garret's instructions [here](https://rstudio-education.github.io/hopr/starting.html) and when you have time read the rest of the book.

`dsir` is not on CRAN and can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("poldham/dsir")
```
## Getting Started

The NCBI API provides higher level access to data if you register with the system and obtain an API token (a unique key). You can use the functions without an API key but access is limited to three calls per second. 

To register for an account with NCBI go [here](https://www.ncbi.nlm.nih.gov/account/register/?back_url=https%3A%2F%2Fwww.ncbi.nlm.nih.gov%2Fhome%2Fdevelop%2Fapi%2F&partners-uri=cms:/account/partners). Note that from June 2021 NCBI will move to third party logins (e.g. using Google, and institutional or government login), see [here](https://ncbiinsights.ncbi.nlm.nih.gov/2021/01/05/important-changes-ncbi-accounts-2021/) for more details. Depending on when you land on this page, follow the latest instructions. 

Inside your account profile you will be able to create an API. Go to your name in the top right corner to access your account details. There you will see a panel called `API Key Management`. Generate an API key and copy it.

The recommended way to store the key is in your R environment with the `usethis` package. As `dsir` is built on `rentrez` we should also install it at the same time

```{r, eval=FALSE}
install.packages("usethis")
install.packages("rentrez")
```

```{r, eval=FALSE}
usethis::edit_r_environ()
```

In the R environment file enter the following, save the file and then `Session > Restart R` so that the change takes effect.

```{r}
ENTREZ_KEY="my Entrez key"
```

When you have restarted R you will be good to go. The approach above is that used by `rentrez` and so you can now happily also use the `rentrez` package without further setup.

### Understanding the databases

NCBI contains a large number of databases. These databases are linked together through a spiders web of connections. 

To understand the databases that are accessible we can use `rentrez`.

```{r}
library(rentrez)
entrez_dbs()
```
For those interested in GenBank (the best known of these databases) note that there is no entry for genbank. What we want here is the `nuccore` (for nucleotice core, also known as nucleotide) database. But, before rushing on, notice all of the other databases such as pubmed for publications, bioproject and biosample. If you work with a public collection you will probably want to take a look at biocollections.

These databases are all linked in various ways and degrees. To see the linkages we use another `rentrez` function.

```{r}
library(rentrez)
rentrez::entrez_db_links("nuccore")
```
We won't go into the detail of these linkages here except to highlight that once you have the data from one database you can then retrieve data from other linked databases. 

### Counting Sequence Records from A Country

Country names appear in NCBI data in two main forms:

1.    In the country field (country tag)
2.    In an organism description (e.g. for a strain or variety)

We want to capture both of these bearing in mind that in a later step we will want to examine and filter out any noisy records. 

To search and obtain country data we follow a four step process:

1. Obtain a count of the records
2. Fetch the records
3. Parse the records to a table (data.frame)
4. Save the data

We want to start by obtaining a count because even though we are working with summary records the data gets big... fast. For example, the summary records for 2.4 million summary records for South Africa take up 45Gb. So, let's not blow up the laptop by accident. 

We use `dsi_count()` to generate counts. Note that the default database for all functions is 'nuccore'.

```{r}
library(dsir)
south_africa <- dsi_count(country = "South Africa", db = "nuccore")
```

We can elaborate on this for other types of count. In international debates on genetic sequence data human genetic material is excluded. Let's see how many of the South Africa records are for humans and how many are non-human.

```{r}
library(dsir)
southafrica_homo <- dsi_count(country = "(South Africa AND Homo sapiens[ORGN])", db = "nuccore")
southafrica_not <- dsi_count(country = "(South Africa NOT Homo sapiens[ORGN])", db = "nuccore")
```

We can figure out if these counts are correct by checking that they add up to the total.

```{r}
(southafrica_homo + southafrica_not) == south_africa
```

Now that we are happy that the counts are working out we can generate other types of count. For example, to avoid dealing with a very large dataset we might want to filter the data by publication date (or optionally record modification date).

We can do this by specifying a range of dates in this case for the first few months until 2021.

```{r}
south_africa_range <- dsi_count(country = "(South Africa) AND 2021/01/01:2021/05/19[PDAT] NOT Homo sapiens[ORGN])", db = "nuccore")
```

When you have registered with NCBI and using an access token the underlying `rentrez` package will take care of any rate limiting when retrieving results. We now turn to retrieving the results. 

### Retrieving results

To retrieve the raw results we use the `dsi_country()` function. In contrast with `dsi_count()` the `dsi_country` function will retrieve all of the results by default. As that can take a number of hours it is good practice to always use `dsi_count()` first. 

```{r, eval=FALSE}
za_sample <- dsi_country(country = "(South Africa) AND 2021/01/01:2021/05/19[PDAT] NOT Homo sapiens[ORGN])", db = "nuccore")
```

In the background the function is making repeated calls to the Entrez API and requesting 10,000 records per time (the maximum). Note that this maximum may vary depending on the NCBI database and it is a good idea to read the [rentrez tutorial](https://cran.r-project.org/web/packages/rentrez/vignettes/rentrez_tutorial.html) on the web history if you want to navigate through these issues. 

When the search has finished we will have a za_sample in the environment (in this case it takes a few minutes). 

The object that comes back is a list that contains chunks of 10,000 records (up to the 36,729 above). These records are in XML format and need to be parsed to a table. 

### Parsing the Results

To parse the results we use `dsi_parse()`. This will iterate over each of the list objects and convert the data to a data.frame that we can save. Note that this can take some time to do. If we simply call dsi_parse the result will be another list (this time of data.frames) so we use `map_df()` from `purrr` (part of the tidyverse) to bind them all together. If you don't have the tidyverse then use `install.packages(tidyverse)` to get it. 

One ppint to note about the parsing of the esummary data is that the existing function does not parse the statistics table inside the summary. The reason for this is that basic statistics such as the sequence length are already included and the stats table does not in reality add much to this. If you want to inspect the stats data to confirm this open up the list object returned above to assess it.

```{r, eval=FALSE}
library(purrr)
za_results <- map_df(za_sample, dsi_parse)
```

As we iterate over the list messages will start to appear. These relate to warnings that will automatically be generated with notes to `Expect warnings`
